---
title: "Cartograflow"
subtitle: "Filtering Matrix for Thematic Flow Mapping"
author: "Françoise Bahoken, Sylvain Blondeau"
date: "`r Sys.Date()`"
output: html_vignette
vignette: >
  \usepackage[ps2pdf,
              bookmarks=true,
              dvipsone=pdftex,                                
              backref,
              ]{hyperref}
  %\VignetteIndexEntry{CartogRaflow}
  %\VignetteEncoding{UTF-8}
  %\SweaveUTF8
  %\VignetteEngine{knitr::rmarkdown}
#editor_options: 
#chunk_output_type: inline
---

`Cartograflow` is designed to filter origin-destination matrix for thematic flow mapping purposes using `{sf}` objects and `{Cartography}` to design the map.

## A - Description of functions

### 1. Preparing flow data sets

#### 1.1 General functions
You can use long "L" or matrix "M" [n*n] flow dataset formats. 

-- `flowtabmat()` is to transform "L" to "M" formats, also to build an empty square matrix from spatial codes.

-- `flowcarre()` is to square a matrix.

-- `flowjointure()` is to performs a spatial join between a flow dataset and a spatial features layer (as a map background) or an external matrix.

-- `flowstructmat()` fixes an unpreviously codes shift in the flow dataset "M" format. If necessary this function is to be used with `flowjointure` and `flowtabmat`.

#### 1.2. Flow computing

-- `flowtype()` is to compute volumn and balance flow from observed flows - respectively to compute symetric and skewmetric matrix from an asymmetric one. 

#### 1.3. Flow reduction

-- `flowreduct()` is to reduce the flow dataset regarding another matrix, e.g. distances travelled. <br/> 

- `metric` is the metric of the distance matrix : _continuous_ (e.g. for meters) or _ordinal_ (e.g. for adjacency). <br/>

If the metric is continuous (e.g for filtering flows by kilometric distances travelled), use:<br/>

- `d.criteria` is for selecting _dmin_ or _dmax_ distance criteria for "continuous" metric ;
Argument `dmin` is for keeping only flows up to a _dmin_ criterion in km ;\cr
Argument `dmax` for selecting values less than a _dmax_ criterion in km.\cr

- `d` is the value of the selected _dmin_ or _dmax_ criteria.

Notice that these arguments can be used as a filter criterion in `flowmap()`.

### 2. Filtering flows

#### 2.1. Filtering flows from concentration analysis 

**Flow concentration analysis:** 

-- `flowgini()` performs a Gini's concentration analysis of the flow features, by computing _Gini coefficient_ and plotting interactive _Lorenz curve_.

To be use before `flowanalysis()`

**Flow filtering according to a concentration criterion:**

-- `flowanalysis()` computes filters criterions based on:

- argument _critflow_ is to filter the flows according to their significativity (% of total of flow information) ; \cr 
- argument _critlink_ is to filter the flows according to their density (% of total features)\cr

These arguments can be used as filter criterion in `flowmap()`.

#### 2.2. Spatial / territorial filtering of flows

**Flow filtering based on a continuous distance criterion**

-- `flowdist()` computes a _continous distance_ matrix from spatial features (area or points). The result is a matrix of the distances travelled between ODs, with flows filtered or not.

**Flow filtering based on an ordinal distance / neighbourhood criterion**:

-- `flowcontig()` compute an _ordinal distance_ matrix from spatial features (area). The result is a matrix of adjacency or k-contiguity of the ODs.

- `background` is the areal spatial features ;\cr
- `code` is the spatial features codes ; \cr
- `k` is to enter the number (k:1,2,...,k) of the contiguity matrix to be constructed : if (k=1), ODs places are adjacent, then the flow have to cross only 1 boundary, else (k=k) ODs places are distant from n borders ;\cr
- `algo` is the algorithm to use for ordinal distance calculation (also Default is "automatic" for "Dijkstra's") ; \cr

Notice that the function automatically returns the maximum (k) number of the spatial layer.

### 3. Flow mapping

-- `flowmap()` is to plot flows as segments or arrows, by acting on the following arguments:\cr 

- `filter` is to filter or not flow's information or features \cr
- `threshold` is used to set the filtering level of the flows when filter= _"True"_ \cr
- `taille` is the value of the width of the flow feature \cr
- `a.head` is the arrow head parameter (in, out, in and out) \cr
- `a.length` is the length of the edges of the arrow head (in inches) \cr
- `a.angle` is the angle from the shaft of the arrow to the edge of the arrow head \cr
- `a.col` is the arrow's color \cr
- `plota` is to add spatial features as map background to the flows's plot \cr
- `add` is to allow to overlay flow features on external spatial features background \cr

## B- Examples of applications

Useful external packages are {dplyr} {sf} {igraph} {rlang} {cartography}.

```{r message=FALSE, include=FALSE}

rm(list=ls())

library(sf)
library(dplyr)
library(cartograflow)
library(cartography)

```

### 1. Load datasets

**Flow dataset**

```{r flowdata_preprocess, echo=FALSE, message=FALSE, warning=FALSE}

# Load example
data<-read.csv2("./data/MOBPRO_ETP.csv",
               header=TRUE,
               sep=";",
               stringsAsFactors=FALSE,
               encoding="UTF-8",
               dec=".",
               check.names=FALSE)

# Variable typing
data$i<-as.character(data$i)
data$j<-as.character(data$j)
data$Fij<-as.numeric(data$Fij)
data$count<-as.numeric(data$count)

str(data)

```

**Select variable and change matrix format**
```{r flowdata_reerse, warning=FALSE, echo=TRUE}

# Selecting useful variables
tabflow<-data%>%select(i,j,Fij)

# Change matrix format (if necessary)
matflow <-flowtabmat(tabflow,matlist="M")
head(matflow[1:4,1:4])
#dim(matflow)

# reverse Change matrix format : from matrix to list
tabflow<-flowtabmat(tab=matflow, matlist="L")
colnames(tabflow)<-c("i","j","Fij")
head(tabflow)

```

**Geographical dataset**
```{r data_preprocess, warning=FALSE, echo=TRUE}

# Load a list of geo codes
ID_CODE<-read.csv2("./data/COD_GEO_EPT.csv",
                   header=TRUE,
                   sep=";",
                   stringsAsFactors=FALSE,
                   encoding="UTF-8",
                   dec=".",
                   check.names=FALSE)
#head(ID_CODE)

CODE<-ID_CODE%>% dplyr::select(COD_GEO_EPT)

colnames(CODE)<-c("CODGEO")
#head(CODE)

```


### 2. Flow computing

*Compute bilateral flows : volum, balance and asymetry*

```{r data_computing, warning=FALSE, echo=TRUE}

# Compute bilateral volum : FSij
matflow_vol<-flowtype(matflow, format="M", "bivolum")

tabflow_vol<-flowtype(tabflow, format="L", "bivolum")

# Compute bilateral balance : FSij
tabflow_net<-flowtype(tabflow, format="L", "bisold")

# Compute all types of bilateral flows, in one 6 columns "L"format matrix
tabflow_all<-flowtype(tabflow, format="L", x="all")
head(tabflow_all) 
 
# Compute flow asymetry
#tabflow_all$FAsy<-(tabflow_all$FDij / tabflow_all$FDij)*100

```

### 3. General Flow mapping

Plot all origin-destination links without any filtering criterion will reveal a graphic complexity ("spaghetti-effect"). So it is better to plot flow up to a simple global paramet (eg. mean).

```{r maps, fig.width=7, message=FALSE, warning=FALSE, include=FALSE}

library(sf)
data(flowdata)
map <- st_read(system.file("shape/MGP_TER.shp", package = "cartograflow"))

```

```{r maps_links, fig.show='hold', fig.width=7, message=TRUE, warning=FALSE, ECHO=FALSE, include=FALSE}

# Add and overlay spatial background 
par(bg = "NA")

knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

# Flowmapping of all links

flowmap(tab=tabflow,
        fij="Fij",
        origin.f = "i",
        destination.f = "j",
        bkg = map,
        code="EPT_NUM",
        nodes.X="X",
        nodes.Y = "Y",
        filter=FALSE,
        add=TRUE
        )

# Map cosmetics
layoutLayer(title = "All origin-destination for commuting in Greater Paris, 2017",
           coltitle ="black",
           author = "Cartograflow, 2020",
           sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
           scale = 2,
           tabtitle = FALSE,
           frame = TRUE,
           col = "grey"
            )
# North arrow
north("topright")

```

**Above-average flowmap** 

```{r maps_flowmean, echo=TRUE, fig.show='hold', fig.width=7, message=FALSE, warning=FALSE, ECHO=FALSE}

library(sf)
data(flowdata)
map <- st_read(system.file("shape/MGP_TER.shp", package = "cartograflow"))

# Add and overlay spatial background 
par(bg = "NA")

knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

# Flow mapping above-average flows
flowmap(tab=tabflow,
        fij="Fij",
        origin.f = "i",
        destination.f = "j",
        bkg = map,
        code="EPT_NUM",
        nodes.X="X",
        nodes.Y = "Y",
        filter=TRUE,
        threshold =(mean(tabflow$Fij)),  #mean value is the level of threshold
        taille=20,           
        a.head = 1,
        a.length = 0.11,
        a.angle = 30,
        a.col="#138913",
        add=TRUE)

# Map Legend
legendPropLines(pos="topleft",
                title.txt="Flows up to 13220 commuters",
                title.cex=0.8,   
                cex=0.5,
                values.cex= 0.7,  
                var=c(mean(tabflow$Fij),max(tabflow$Fij)), 
                lwd=5, 
                frame = FALSE,
                col="#138913",
                values.rnd = 0
                )

#Map cosmetic

layoutLayer(title = "Commuters up to above-average in Greater Paris",
           coltitle ="black",
           author = "Cartograflow, 2020",
           sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
           scale = 2,
           tabtitle = FALSE,
           frame = TRUE,
           col = "grey"
            )

# North arrow
north("topright")

```

### 4. Concentration analysis for Flow mapping

**Main functions**

`flowgini()` and `flowanalysis()`

**Compute the concentration of flow values**

```{r analysis, fig.show='hold',message = TRUE , warning=FALSE}

tabgini<-flowgini(ODpts = tabflow,
                  origin="i",destination = "j",valflow = "Fij",
                  lorenz.plot = FALSE)

# Interpretation ; The flows are quite concentrated on a few links, the Gini coefficent is equal to 73.16% 

```

**Plot the corresponding interactive Lorenz' curve**

```{r analysis_lorenz, fig.show='hold',message = TRUE , warning=FALSE}

head(tabgini)

flowgini(ODpts = tabflow,
         origin="i",destination = "j",valflow = "Fij",
         lorenz.plot = TRUE)

```

**Compute the "critflow" parameter (ex. significance)**

```{r echo=TRUE, fig.show='hold', message=TRUE, warning=FALSE}

flowanalysis(tabgini,
             critflow = 0.8,
             result = "signif")

# Interpretation : Flow values up to 13442 are the 80% largest one corresponding to 23,14 % of the total links' features.

```

**Flowmap filtered according to flows values significance**

Using the `flowanalysis()` _"critflow"_ value to select flows.

```{r echo=TRUE, fig.show='hold', message=TRUE, warning=FALSE}

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

# For mapping flow up to 13342

flowmap(tab=tabflow,
        fij="Fij",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        add=TRUE,
        filter=TRUE,
        threshold=13442,    
        taille=15,           
        a.head = 1,
        a.length = 0.11,
        a.angle = 30,
        a.col="#138913")

# Map Legend
legendPropLines(pos="topleft",
                title.txt="Commuters up to 13442\n (80% of the largest flows)",
                title.cex=0.8,   
                cex=0.5,
                values.cex= 0.7,  
                var=c(13442,max(tabflow$Fij)), 
                lwd=15, 
                frame = FALSE,
                col="#138913",
                values.rnd = 0
                )

#Map cosmetic

layoutLayer(title = "Significant professional mobility in Greater Paris",
           coltitle ="black",
           author = "Cartograflow, 2020",
           sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
           scale = 2,
           tabtitle = FALSE,
           frame = TRUE,
           col = "grey",
            )
# north arrow
north("topright")

```

**Flowmap filtered according to flow features' density**

Using the `flowanalysis()` _"critlink value"_ to select flows and then flowmap.

```{r fig.show='hold', warning=FALSE, include=FALSE}

flowanalysis(tabgini,
             critlink = 0.1,
             result = "density")

# Interpretation : Flows up to 45772 are the 58.12% largest one corresponding to 10 % of the total links

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

# For mapping 10% of the total features as flows up to 45772 commuters

flowmap(tab=tabgini,
        fij="Fij",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        add=TRUE,
        plota = TRUE,
        filter=TRUE,
        threshold=45772,    
        taille=15,           
        a.head = 1,
        a.length = 0.15,
        a.angle = 30,
        a.col="#138913")

# Map Legend
legendPropLines(pos="topleft",
                title.txt="Number of commuters up to 45772\n(10% of the links)",
                title.cex=0.8,   
                cex=0.5,
                values.cex= 0.7,  
                var=c(45772,max(tabgini$Fij)), 
                lwd=15, 
                frame = FALSE,
                col="#138913",
                values.rnd = 0
                )

#Map cosmetic
layoutLayer(title = "Low density professional mobility in Greater Paris",
           coltitle ="black",
           author = "Cartograflow, 2020",
           sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
           scale = 2,
           tabtitle = FALSE,
           frame = TRUE,
           col = "grey"
            )

# north arrow
north("topright")

```

### 5. Filtering flows by continous distance travelled

Filtering an Origin-Destination matrix with a **continuous distance matrix** (in kilometers). The aim is to plot flow 1) less than a maximum distance value, 2) above a minimum distance travelled criterion or 3) on a range of distances.

**Main function**

`flowdist()`

**Useful additional functions**

`flowjointure()` and `flowreduct()` with the `metric` parameter:" _continuous_".

**Compute distance matrix**

Function aims first to compute a distance matrix then to reduce the matrix and finally to plot the (filtered) flows. Example is for euclidian distance.

```{r echo=TRUE, fig.show='hold'}

head(tabflow)

tab<-flowjointure(geom="area",DF.flow=tabflow,origin = "i",destination = "j",
                   bkg=map,id="EPT_NUM",x="X",y="Y")

tab.distance<-flowdist(tab,
                       dist.method = "euclidian",
                       result = "dist")

tab.distance<-tab.distance %>% select(i,j,distance)
tab<-tab %>% select(i,j,ydata)
head(tab.distance)

```

**Flow reduction according to a maximum distance travelled**

Using for mapping flow less than the maximum distance travelled criterion.

```{r echo=TRUE, fig.show='hold'}

#reduce the flow dataset from a selected distance travelled (eg. 8.5 km)
library(rlang)

tab.flow<-flowreduct(tab,
                     tab.distance,
                     metric = "continous",
                     d.criteria = "dmax", #max distance parameter 
                     d = 8567)        #max distance value - Q1 : 8567 km

#select for all i,j flow values up to 0
flow.d<-tab.flow %>%
        select(i,j,flowfilter) %>%
        filter(flowfilter !=0)
head(flow.d)

```

**Flowmap filtered according to a maximum distance travelled parameter** 

Using the `flowreduct()` *d.criteria as _"dmax"_ distance parameter* to plot flows less than the maximum distance criterion (here : 8,5 km).

```{r echo=TRUE, fig.show='hold'}

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

#Flowmap : flow travelled less than 8.5 km  (as the first quartile Q1)

flowmap(tab=flow.d,
        fij="flowfilter",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=TRUE,
        taille=8,           
        a.head = 1,
        a.length = 0.11,
        a.col="#f7714f",
        add=TRUE)

#Map legend
legendPropLines(pos="topleft",
                title.txt="Number of commuters\n(distance travelled less than 8,5 km)",
                title.cex=0.8,    
                cex=0.5,
                values.cex= 0.7,  
                var=c(min(flow.d$flowfilter),8567), 
                col="#f7714f",
                lwd=8,
                frame = FALSE,
                values.rnd = 0
                )

#Map cosmetic
layoutLayer(title = "Professional mobility in Greater Paris : short distance travelled",
            author = "Cartograflow, 2020",
            sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
            scale = 2,
            tabtitle = FALSE,
            frame = TRUE,
            col = "grey",
            coltitle ="black"
            )

# north arrow
north("topright")

```

**Flowmap filtered according to a minimum distance travelled parameter**

Using the `flowreduct()` *d.criteria as _"dmin"_ distance parameter* to plot flows above the minimum distance criterion (here : 20 km).

```{r fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}

#head(tabflow)

tab<-flowjointure(geom="area",DF.flow=tabflow,origin = "i",destination = "j",
                   bkg=map,id="EPT_NUM",x="X",y="Y")

tab.distance<-flowdist(tab,
                       dist.method = "euclidian",
                       result = "dist")

tab.distance<-tab.distance %>% select(i,j,distance)
tab<-tab %>% select(i,j,ydata)
head(tab.distance)

tab.flow<-flowreduct(tab,
                     tab.distance,
                     metric = "continous",
                     d.criteria = "dmin",  
                     d = 19234)        #Q2 : 14518 km - Q3:19234 km

#select for all i,j flow values above to 0
flow.d<-tab.flow%>%
       select(i,j,flowfilter)%>%
        filter(flowfilter !=0)

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

#Flowmap : flow travelled up to 20 km (as the third quartile Q3)

par(mar=c(0,0,1,0))

extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

flowmap(tab=flow.d,
        fij="flowfilter",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=TRUE,
        taille=8,           
        a.head = 1,
        a.length = 0.11,
        a.col="#f7714f",
        add=TRUE)

# Map legend
legendPropLines(pos="topleft",
                title.txt="Number of commuters\n(distance travelled more than 19 km)",
                title.cex=0.8,    
                cex=0.5,
                values.cex= 0.7,  
                var=c(19234, max(flow.d$flowfilter)), 
                col="#f7714f",
                lwd=8, 
                frame = FALSE,
                values.rnd = 0
                )
# Map cosmetics
layoutLayer(title = "Professional mobility in Greater Paris : long distance travelled",
            author = "Cartograflow, 2020",
            sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
            scale = 2,
            tabtitle = FALSE,
            frame = TRUE,
            col = "grey",
            coltitle ="black")

# north arrow
north("topright")

```

**Flowmap reduction and filtering according to distance matrix**
------------------------------

Using the `flowreduct()` *d.criteria as "dmax" distance parameter** in association with a `flowmap()` **_"threshold"_** value to map flows between two values (here between the first and third quartiles [Q1 - Q3]) - Not show.

```{r fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}

#head(tabflow)

tab<-flowjointure(geom="area",DF.flow=tabflow,origin = "i",destination = "j",
                   bkg=map,id="EPT_NUM",x="X",y="Y")

tab.distance<-flowdist(tab,
                       dist.method = "euclidian",
                       result = "dist")

tab.distance<-tab.distance %>% select(i,j,distance)
tab<-tab %>% select(i,j,ydata)
head(tab.distance)

tab.flow<-flowreduct(tab,
                     tab.distance,
                     metric = "continous",
                     d.criteria = "dmax",  
                     d = 19234)       #Q3:19234 km

#select for all i,j flow values up to 8567  
flow.d<-tab.flow%>%
       select(i,j,flowfilter)%>%
        filter(flowfilter >8567)      #Q1=8567km

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

#Flowmap : flow travelled up to (Q3)

flowmap(tab=flow.d,
        fij="flowfilter",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=TRUE,
        taille=8,           
        a.head = 1,
        a.length = 0.11,
        a.col="#f7714f",
        add=TRUE)

# Map legend
legendPropLines(pos="topleft",
                title.txt="Number of commuters\n(distance travelled between 8,5 and 19 km)",
                title.cex=0.8,    
                cex=0.5,
                values.cex= 0.7,  
                var=c(8567, max(flow.d$flowfilter)), 
                col="#f7714f",
                lwd=8, 
                frame = FALSE,
                values.rnd = 0
                )

# Map cosmetic
layoutLayer(title = "Professional mobility in Greater Paris : median distance travelled",
            author = "Cartograflow, 2020",
            sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
            scale = 5,
            tabtitle = TRUE,
            frame = TRUE,
            col = "grey",
            coltitle ="black")
```

### 6. Filtering flows by ordinal distance travelled

Filtering an Origin-Destination matrix with an **ordinal distance matrix**. This matrix describes a neighborhood space defined by a number (k) of boundaries to be crossed in order to reach a place of destination from a place of origin. The aim is to map local flows that are either adjacent or located at low (k).

**Main function**

`flowcontig()`

**Useful additional function**

`flowreduct()` with the `metric` parameter: _"ordinal"_.

**Computes the neighbouring graph**

Example is for neighbouring areas which share a common boundary (k=1) 

```{r echo=TRUE, include=TRUE, message=FALSE, fig.show='hold', warning=FALSE}

library(igraph)
## Neighbouring graph (order k= 1)
graph_ckij_1<-flowcontig(bkg=map, code="EPT_NUM", 
                         k=1, algo = "automatic")
#Max order = "3"
```

Plot the neighbouring graph (k=1)
```{r echo=FALSE, include=FALSE, message=FALSE, fig.show='hold', warning=FALSE}

flowmap(tab=graph_ckij_1,
        fij="ordre",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=FALSE)

mtext("Study Area Neighbourhood Plot (k=1)", side=3)

```

**Flow reduction according to the neighbouring graph**

Reducing flow matrix by the neighbouring graph (k=1)

```{r echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE}
library(rlang)

#head(tabflow)
#head(graph_ckij_1)

reduc_k1<-flowreduct(tabflow,
                  graph_ckij_1,
                  metric = "ordinal")
head(reduc_k1)
```

**Flowmap between adjacent areas** 

Using (k=1) parameter

```{r echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE}

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

#Flowmap : flow travelled between adjacent areas

flowmap(tab=reduc_k1,
        fij="flow",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=TRUE,
        taille=8,
        a.head = 1,
        a.length = 0.11,
        a.col="#0e7fe3",
        add=TRUE
        )

# Map Legend
legendPropLines(pos="topleft",
                title.txt="Number of commuters in adjacent places\n(k=1)",
                title.cex=0.8,    
                cex=0.5,
                values.cex= 0.7,  
                var=c(min(reduc_k1$flow),max(reduc_k1$flow)), 
                col="#0e7fe3",
                lwd=8, 
                frame = FALSE,
                values.rnd = 0
                )

# Map cosmetic
layoutLayer(title = "Professional mobility in Greater Paris between 1-neighbouring municipalities",
            author = "Cartograflow, 2020",
            sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
            scale = 2,
            tabtitle = FALSE,
            frame = TRUE,
            col = "grey",
            coltitle ="black")

```

**Flowmap between non adjacent areas** 

Using for example (k=3) parameter.

```{r fig.show='hold', message=FALSE, warning=FALSE, include=FALSE}

library(igraph)

## Neighbouring graph (Max(k)=3)
graph_ckij_3<-flowcontig(bkg=map, code="EPT_NUM",k=3)

# Flow reduction
#head(tabflow)

reduc_k3<-flowreduct(tabflow,
                  graph_ckij_3,
                  metric = "ordinal")
#head(reduc_k3)

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

#Flowmap : flow travelled between non adjacent areas

flowmap(tab=reduc_k3,
        fij="flow",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=TRUE,
        taille=8,
        a.head = 1,
        a.length = 0.11,
        a.col="#0e7fe3",
        add=TRUE 
        )

# Map Legend
legendPropLines(pos="topleft",
                title.txt="Number of commuters between non adjacent places\n(k=3)",
                title.cex=0.8,    
                cex=0.5,
                values.cex= 0.7,  
                var=c(min(reduc_k3$flow),max(reduc_k3$flow)), 
                col="#0e7fe3",
                lwd=8, 
                frame = FALSE,
                values.rnd = 0
                )

# Map cosmetic
layoutLayer(title = "Professional mobility in Greater Paris between 3-neighbouring municipalities",
            author = "Cartograflow, 2020",
            sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
            scale = 2,
            tabtitle = FALSE,
            frame = TRUE,
            col = "grey",
            coltitle ="black")

```

**Flowmap reducing and filtering according to ordinal matrix**

Using the `flowcontig()` **(k)** parameter in association to the `flowmap()` **threshold** parameter to map above-average flows that occur between adjacent areas. - Not show.

```{r flows_criteria, echo=TRUE, fig.show='hold', fig.width=7, message=FALSE, warning=FALSE, ECHO=FALSE}

#Computes k=1
library(igraph)

#Neighbouring graph k=1
graph_ckij_1<-flowcontig(bkg=map, code="EPT_NUM",k=1)

#Flow reduction
reduc_k1<-flowreduct(tabflow,
                  graph_ckij_1,
                  metric = "ordinal")

#Mean flow value
reduc_k1_mean<-mean(reduc_k1$flow)
mean<-mean(tabflow$Fij)

# mean value of reduc_k1_mean =18591
# mean value of tabflow =13220

```

```{r echo=TRUE, fig.show='hold', message=FALSE, warning=FALSE}

# Graphic parameters
knitr::opts_chunk$set(fig.width=6, fig.height=6)
par(mar=c(0,0,1,0))
extent <- c(2800000, 1340000, 6400000, 4800000)
resolution<-150

# Overlay a spatial background 
par(bg = "NA")

# Add the corresponding background 
plot(st_geometry(map), col = NA, border=NA, bg="#dfe6e1")
plot(st_geometry(map), col = "light grey", add=TRUE)

#Flowmap : flow travelled between adjacent areas

flowmap(tab=reduc_k1,
        fij="flow",origin.f = "i",destination.f = "j",
        bkg = map,code="EPT_NUM",nodes.X="X",nodes.Y = "Y",
        filter=TRUE,
        threshold = 13220, #reduc_k1 mean value is 18591 ; tabflow mean value is 13220
        taille=8,
        a.head = 1,
        a.length = 0.11,
        a.col="#0e7fe3",
        add=TRUE 
        )

# Map Legend
legendPropLines(pos="topleft",
                title.txt="Number of above-average flows that occur between adjacent areas\n(k=1)",
                title.cex=0.8,    
                cex=0.5,
                values.cex= 0.7,  
                var=c(min(reduc_k1$flow),max(reduc_k1$flow)), 
                col="#0e7fe3",
                lwd=8, 
                frame = FALSE,
                values.rnd = 0
                )

# Map cosmetic
layoutLayer(title = "Professional mobility in Greater Paris between 1-neighbouring municipalities",
            author = "Cartograflow, 2020",
            sources = "Data : INSEE, 2017 ; Basemap : APUR, RIATE, 2018.",
            scale = 2,
            tabtitle = FALSE,
            frame = TRUE,
            col = "grey",
            coltitle ="black")

```


## C- Sample datasets

-- _Statistical dataset_ : <br/>
- INSEE - Base flux de mobilité (2015) 
- URL : https://www.insee.fr/fr/statistiques/fichier/3566008/rp2015_mobpro_txt.zip

-- _Geographical dataset_ :
- municipalities : IGN, GEOFLA 2015 v2.1 
- Greater Paris : APUR, UMS 2414 RIATE, 2018.

## Reproducibility

```{r lecho=TRUE, fig.show='hold'}

sessionInfo()

```
